Combining network-guided GWAS to discover susceptibility mechanisms for breast cancer

### 2020-04-20 manuscript/manuscript.tex
feedback on stability network

\begin{abstract}
Systems biology provides a comprehensive approach to biomarker discovery and biological hypothesis building. Indeed, it allows to jointly consider the statistical association between gene variation and a phenotype, and the biological context of each gene, represented as a network. In this work, we study the utility of six methods that identify subnetworks with higher association scores to a phenotype to discover new biomarkers for breast cancer susceptibility. We interrogate a familial breast cancer genome-wide association study (GWAS) focused on \emph{BRCA1/2} negative French women. We perform an in-depth benchmarking of the methods with regards to size of the solution subnetwork, their utility as biomarkers, and the stability and the runtime of the methods. By trading statistical stringency for biological meaningfulness, most network methods give more compelling results than standard SNP- and gene-level analyses, recovering causal subnetworks tightly related to cancer susceptibility. Interestingly, a combination of solution subnetworks provided a concise subnetwork of 93 genes, enriched in known breast cancer susceptibility genes (\emph{BABAM1}, \emph{BLM}, \emph{CASP8}, \emph{FGFR2}, and \emph{TOX3}, Fisher's exact test P-value~=~$7.8 \times 10^{-5}$) and occupying more central positions in the network than average. Additionally, it shows a general alteration of the neighborhood of \emph{COPS5}, a gene related to multiple hallmarks of cancer. Finally, we find a significantly large overlap between the genes in the consensus network and the genes significantly associated in the largest GWAS on susceptibility to breast cancer. In conclusion, this article shows the pertinence of network-based analyses to tackle known issues with GWAS, namely lack of statistical power and of interpretable solutions. Project-agnostic implementations of each of the network methods are available at \url{https://github.com/hclimente/gwas-tools} to facilitate their application to other GWAS datasets.
\end{abstract}

@@

Most of the network methods, even the consensus, were highly unstable, raising questions about the veracity of the results. We built a consensus network using the 30 solutions from the 6 different methods on 5 different splits containing 80\% of the data (Section \ref{methods:comparison}). Such network would capture the subnetworks more often found altered, and hence should be more resistant to noise. We used only genes selected in at least 7 of the settings (percentile 99\%). The resulting stablility-based consensus was composed of 68 genes. This network shares most of the proterties of the consensus: breast cancer susceptibility genes are overrepresented (P-value = 3 \texttimes{} 10\textsuperscript{-4}), as well as genes involved in mitochondrial translation and the attenuation phase (adjusted P-values 0.001 and 3 \texttimes{} 10\textsuperscript{-5} respectively), the selected genes are more central than average (P-value = 1.1 \texttimes{} 10\textsuperscript{-14}), and 19 of them are isolated.

However, the shared global properties of the network, the lack of stability results in different genes being selected. In this case, the most central gene of the network is \emph{CUL3}, which is absent from the consensus network. Indeed, \emph{CUL3} has a low association score in both GENESIS and BCAC (P-values of 0.04 and 0.26, respectively). \emph{CUL3} is a component of Cullin-RING ubiquitin ligases. It impacts the protein levels of multiple genes relevant for cancer progression \cite{Chen2016}, and its overexpression was also linked to increased sensitivity to carcinogens \cite{Loignon2009}.

@@

In recent years, the ability of GWAS to unravel the mechanisms leading to complex diseases has been put into questions \cite{boyle_expanded_2017}. The omnigenic model proposes that gene functions are interwoven with each other in a dense co-function network. The practical consequence is that larger and larger GWAS will lead to the discovery an uninformative wide-spread pleiotropy. Other than such problems, discovery in GWAS is hindered by a conservative statistical framework. Network methods tackle both issues by using both the association score and an interaction networks to consider the biological context of each of the genes and SNPs. Based on what could be considered diverse interpretations of the omnigenic model, several methods for network-guided biomarker discovery have been proposed in recent years. In this article we evaluate the relevance of six network methods by examining a GWAS dataset on familial breast cancer.

Most of these methods produced a relevant subset of biomarkers, recapitulating known breast cancer susceptibility genes. On median, the selected genes and SNPs were more central than average, in accordance with the observation that disease genes are more relatively central \cite{pinero_uncovering_2016}. However, very central nodes are also more likely to be connecting a given random pair of nodes, making them more likely to be selected by these network methods. Hence, further work is needed characterize the impact of centrality on these methods' outputs. Despite their similarities, the solutions were notably different. At one end of the spectrum, SConES and heinz preferred small, highly associated solutions, providing a conveniently short list of biomarkers, at the expense of not shedding much light on the etiology of the disease. On the other end, SigMod and dmGWAS gravitate towards larger, less associated solutions which provide a wide overview of the biological context. While this deepens our understanding of the disease and provide biological hypotheses, they require further analyses, which might deter unexperienced practitioners. HotNet2 balances both approaches at the expense of producing the largest solution: a constellation of many, highly associated, small subnetworks. Despite these differences, all solutions are equally bad at discriminating cases from controls. In addition, all methods are remarkably unstable, yielding different solutions for slightly different inputs. This might partly be caused by the instability of the P-values themselves in low statistical power settings \cite{halseyFickleValueGenerates2015}. Hence, heinz's conservative transformation of P-values, which favors only the most extreme ones, leads to improved stability. Another source of instability might be the redundancy inherent to biological networks. We built a stable consensus, which accounts for that instability. Encouragingly, this network shared the same biological and topological properties as the individual solutions, despite the few genes in common due to instability. Overall, this supports that instability might be caused by redundant but equivalent biological mechanisms, and hence it validates the conclusions obtained on the individual solutions and the consensus.
\cazcom{}{I think we need to discuss this. Why aren't we able to classify cases from controls (multiple non-genetic factors, disease heterogeneity...)? Why is our approach still useful if we aren't, in fact, discovering biomarkers? (it points at relevant pathways, etc) }

@@

As not all databases compile the same interactions, the choice of the PPIN determines the final output. In this work we used exclusively interactions from HINT from high-throughput experiments. This responds to concerns about adding interactions identified in targeted studies and prone to a ``rich getting richer'' phenomenon: popular genes have a higher proportion of their interactions described \cite{cai_broker_2010,das_hint:_2012}. Their presense might bias the results of the presented methods, as they change the topology of the network by adding nodes closer to other nodes than average. On the other hand, \citet{huang_systematic_2018} found that the best predictor of the performance of a network for disease gene discovery is the size of the network, which supports using the largest amount of interactions. When we compared the impact of using a larger network containing interactions from both high-throughput experiment and the literature (Section~\ref{methods:networks}), we found that for most of the methods it did not greatly change the size or the stability of the solution, the classification accuracy, or the runtime (Supplementary Figure~\ref{sfig:lc_ht_comparison}). This supports using only interactions from high-throughput experiments, which produces apparently similar solutions and avoids falling into ``circular reasonings'', where the best known genes are artificially pushed into the selected solutions.

>>>>

\begin{abstract}
Systems biology provides a comprehensive approach to biomarker discovery and biological hypothesis building. Indeed, it allows to jointly consider the statistical association between gene variation and a phenotype, and the biological context of each gene, represented as a network. In this work, we study the utility of six methods that identify subnetworks with higher association scores to a phenotype to discover new biomarkers for breast cancer susceptibility. We interrogate a familial breast cancer genome-wide association study (GWAS) focused on \emph{BRCA1/2} negative French women. We perform an in-depth benchmarking of the methods with regards to size of the solution subnetwork, their utility as biomarkers, and the stability and the runtime of the methods. By trading statistical stringency for biological meaningfulness, most network methods give more compelling results than standard SNP- and gene-level analyses, recovering causal subnetworks tightly related to cancer susceptibility. Interestingly, a combination of solution subnetworks provided a concise subnetwork of 93 genes, enriched in known breast cancer susceptibility genes (\emph{BABAM1}, \emph{BLM}, \emph{CASP8}, \emph{FGFR2}, and \emph{TOX3}, Fisher's exact test P-value~=~$7.8 \times 10^{-5}$) and occupying more central positions in the network than average. \cazcom{Additionally, it shows a general alteration of the neighborhood of \emph{COPS5}, a gene related to multiple hallmarks of cancer}{Update to talk about the stable consensus and \textit{CUL3} instead? If we find it more reliable?}. Finally, we find a significantly large overlap between the genes in the consensus network and the genes significantly associated in the largest GWAS on susceptibility to breast cancer. In conclusion, this article shows the pertinence of network-based analyses to tackle known issues with GWAS, namely lack of statistical power and of interpretable solutions. Project-agnostic implementations of each of the network methods are available at \url{https://github.com/hclimente/gwas-tools} to facilitate their application to other GWAS datasets.
\end{abstract}

@@

Most of the network methods, even the consensus, were highly unstable, raising questions about the \caz{veracity}{reliability} of the results. We built a \caz{}{new, stable} consensus network using the 30 solutions \caz{from}{obtain by running} the 6 different methods on 5 different splits \caz{}{each} containing 80\% of the data (Section \ref{methods:comparison}). Such a network \caz{would}{is expected to} capture the subnetworks more often found altered, and hence should be more resistant to noise. We used only genes selected in at least 7 of the settings (\cazcom{percentile 99\%}{I'm reluctant to talk about percentiles here, as the genes aren't ranked with each a different value -- I'd rather say ``corresponding to selecting 1\% of the genes'' or something like that}). The resulting stablility-based consensus was composed of 68 genes \cazcom{}{Figure??? Also add  a figure with all gene names in the supplementaries, like Supp Fig 4?}. This network shares most of the proterties of the consensus: breast cancer susceptibility genes are overrepresented (P-value = 3 \texttimes{} 10\textsuperscript{-4}), as well as genes involved in mitochondrial translation and the attenuation phase (adjusted P-values 0.001 and 3 \texttimes{} 10\textsuperscript{-5} respectively); the selected genes are more central than average (P-value = 1.1 \texttimes{} 10\textsuperscript{-14}); and \cazcom{19 of them are isolated}{what property of the previous consensus network is this?}.

However, \caz{the shared global properties of the network}{although this new network exhibits similar global properties as the previous one}, the lack of stability results in different genes being selected. In this case, the most central gene of the network is \emph{CUL3}, which is absent from the \caz{}{previous} consensus network. Indeed, \emph{CUL3} has a low association score in both GENESIS and BCAC (P-values of 0.04 and 0.26, respectively). \emph{CUL3} is a component of Cullin-RING ubiquitin ligases. It impacts the protein levels of multiple genes relevant for cancer progression \cite{Chen2016}, and its overexpression was also linked to increased sensitivity to carcinogens \cite{Loignon2009}.

@@

In recent years, the ability of GWAS to unravel the mechanisms leading to complex diseases has been called into question \cite{boyle_expanded_2017}. The omnigenic model proposes that gene functions are interwoven with each other in a dense co-function network. The practical consequence is that larger and larger GWAS will lead to the discovery of an uninformative wide-spread pleiotropy. \caz{Other than such problems}{In addition}, discovery in GWAS is hindered by a conservative statistical framework. Network methods tackle \caz{both}{these two} issues by using both the association score and an interaction network to \caz{consider}{take into consideration} the biological context of each of the genes and SNPs. Based on what could be considered diverse interpretations of the omnigenic model, several methods for network-guided biomarker discovery have been proposed in recent years. In this article we evaluate the relevance of six network methods by examining a GWAS dataset on familial breast cancer.

Most of these methods produced a relevant subset of biomarkers, recapitulating known breast cancer susceptibility genes. \caz{On median}{In general} \cazcom{}{I don't think ``On median'' exists, unfortunately}, the selected genes and SNPs were more central than average, in accordance with the observation that disease genes are more relatively central \cite{pinero_uncovering_2016}. However, very central nodes are also more likely to be connecting a given random pair of nodes, making them more likely to be selected by these network methods. Hence, further work is needed to characterize the impact of centrality on these methods' outputs. Despite their similarities, the solutions were notably different. At one end of the spectrum, SConES and heinz preferred small, highly associated solutions, providing a conveniently short list of biomarkers, at the expense of not shedding much light on the etiology of the disease. On the other end, SigMod and dmGWAS gravitate towards larger, less associated solutions which provide a wide overview of the biological context. While this deepens our understanding of the disease and provide biological hypotheses, they require further analyses, which might deter unexperienced practitioners. HotNet2 balances both approaches at the expense of producing the largest solution: a constellation of many, highly associated, small subnetworks. Despite these differences, all solutions are equally bad at discriminating cases from controls. In addition, all methods are remarkably unstable, yielding different solutions for slightly different inputs. This might partly be caused by the instability of the P-values themselves in low statistical power settings \cite{halseyFickleValueGenerates2015}. Hence, heinz's conservative transformation of P-values, which favors only the most extreme ones, leads to improved stability. Another source of instability might be the redundancy inherent to biological networks. \cazcom{We built a stable consensus, which accounts for that instability. Encouragingly, this network shared the same biological and topological properties as the individual solutions, despite the few genes in common due to instability. Overall, this supports that instability might be caused by redundant but equivalent biological mechanisms, and hence it validates the conclusions obtained on the individual solutions and the consensus.}{This comes too early, the next paragraph discusses why and how we built the consensus network and this bit should only arrive afterwards}
\cazcom{}{I think we need to discuss this. Why aren't we able to classify cases from controls (multiple non-genetic factors, disease heterogeneity...)? Why is our approach still useful if we aren't, in fact, discovering biomarkers? (it points at relevant pathways, etc) }

@@

As not all databases compile the same interactions, the choice of the PPIN determines the final output. In this work we used exclusively interactions from HINT from high-throughput experiments. This responds to concerns about adding interactions identified in targeted studies and prone to a ``rich getting richer'' phenomenon: popular genes have a higher proportion of their interactions described \cite{cai_broker_2010,das_hint:_2012}. Their presence might bias the results of the presented methods, as they change the topology of the network by adding nodes closer to other nodes than average. On the other hand, \citet{huang_systematic_2018} found that the best predictor of the performance of a network for disease gene discovery is the size of the network, which supports using the largest amount of interactions. When we compared the impact of using a larger network containing interactions from both high-throughput experiment and the literature (Section~\ref{methods:networks}), we found that for most of the methods it did not greatly change the size or the stability of the solution, the classification accuracy, or the runtime (Supplementary Figure~\ref{sfig:lc_ht_comparison}). This supports using only interactions from high-throughput experiments, which produces apparently similar solutions and avoids falling into ``circular reasonings'', where the best known genes are artificially pushed into the selected solutions.
